---
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)

set.seed(945)

# Random generate two Xs with no correlation
x1 <- rnorm(1e4)
x2 <- rnorm(1e4)

# Plot and get no correlation
ggplot(tibble(x1, x2), aes(x1, x2)) +
  geom_point(alpha = 0.1)


y <- rnorm(1e3, mean = 1 + 3 * x1 + 2 * x2, sd = 0.1)


```


```{r}

# Plot and get no correlation
ggplot(tibble(x1, y), aes(x1, y)) +
  geom_point(alpha = 0.1)

summary(lm(y ~ x1))

summary(lm(y ~ x1 + x2))

# 
```

if we have two predictors that are unrelated, you are still going to 
if our predictors are correlated with eachother then our estimate for X1 coefficient is the combined effect of both x1 and x2. And that's why our coefficient is biased

if x 1 and x2 are both influencing y but are unrelated (NO BIAS)
but if x1 and x2 are correlated (you are going to get a biased)

KEY TAKEAWAY:
If you leave out a variable you are at risk of causing a variable bias. In order for that to happen your variables have to be correlated to each other. 
